---
title: "DSC520 Week 02 Assignment"
author: "Kurt Stoneburner"
date: "6/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(pastecs)

#readDir <- "C:\\Users\\newcomb\\DSCProjects\\dsc520_github"
readDir <- "L:\\stonk\\projects\\DSC\\dsc520"
survey_df <- read.csv(paste(readDir,"\\data\\acs-14-1yr-s0201.csv",sep=""))

# 1. What are the elements in your data (including the categories and data types)?
#   
#   2. Please provide the output from the following functions: str(); nrow(); ncol()
# 
# 3. Create a Histogram of the HSDegree variable using the ggplot2 package.
# 
# a. Set a bin size for the Histogram.
# 
# b. Include a Title and appropriate X/Y axis labels on your Histogram Plot.
# 
# 4. Answer the following questions based on the Histogram produced:
#   
#   a. Based on what you see in this histogram, is the data distribution unimodal?
#   
#   b. Is it approximately symmetrical?
#   
#   c. Is it approximately bell-shaped?
#   
#   d. Is it approximately normal?
#   
#   e. If not normal, is the distribution skewed? If so, in which direction?
#   
#   f. Include a normal curve to the Histogram that you plotted.
# 
# g. Explain whether a normal distribution can accurately be used as a model for this data.
# 
# 5. Create a Probability Plot of the HSDegree variable.
# 
# 6. Answer the following questions based on the Probability Plot:
#   
#   a. Based on what you see in this probability plot, is the distribution approximately normal? Explain how you know.
# 
# b. If not normal, is the distribution skewed? If so, in which direction? Explain how you know.
# 
# 7. Now that you have looked at this data visually for normality, you will now quantify normality with numbers using the stat.desc() function. Include a screen capture of the results produced.
# 
# 8. In several sentences provide an explanation of the result produced for skew, kurtosis, and z-scores. In addition, explain how a change in the sample size may change your explanation?
```
What are the elements in your data (including the categories and data types)

1. There are 8 elements in the dataset
  - Id              : Categorical, used as an identifier
  - Id2             : Categorical, used as an identifier
  - Geography       : Categorical
  - PopGroupID      : Categorical
  - RacesReported   : Quantitative
  - HSDegree        : Quantitative
  - BachDegree      : Quantitative

2. 
&nbsp;&nbsp;&nbsp;&nbsp;**head(survey_df)**

```{r head, echo=FALSE}
head(survey_df)
```
&nbsp;&nbsp;&nbsp;&nbsp;**str(survey_df)**
```{r str, echo=FALSE}
str(survey_df)
```
&nbsp;&nbsp;&nbsp;&nbsp;**nrow(survey_df)**
``` {r nrow, echo=FALSE}  
nrow(survey_df)
```
&nbsp;&nbsp;&nbsp;&nbsp;**ncol(survey_df)**
``` {r ncol, echo=FALSE}  
ncol(survey_df)
```
3.

4.

5.

6.

7.



