---
title: "DSC520 Week 02 Assignment"
author: "Kurt Stoneburner"
date: "6/11/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(pastecs)

readDir <- "C:\\Users\\newcomb\\DSCProjects\\dsc520_github"
#readDir <- "L:\\stonk\\projects\\DSC\\dsc520"
survey_df <- read.csv(paste(readDir,"\\data\\acs-14-1yr-s0201.csv",sep=""))

# 1. What are the elements in your data (including the categories and data types)?
#   
#   2. Please provide the output from the following functions: str(); nrow(); ncol()
# 
# 3. Create a Histogram of the HSDegree variable using the ggplot2 package.
# 
# a. Set a bin size for the Histogram.
# 
# b. Include a Title and appropriate X/Y axis labels on your Histogram Plot.
# 
# 4. Answer the following questions based on the Histogram produced:
#   
#   a. Based on what you see in this histogram, is the data distribution unimodal?
#   
#   b. Is it approximately symmetrical?
#   
#   c. Is it approximately bell-shaped?
#   
#   d. Is it approximately normal?
#   
#   e. If not normal, is the distribution skewed? If so, in which direction?
#   
#   f. Include a normal curve to the Histogram that you plotted.
# 
# g. Explain whether a normal distribution can accurately be used as a model for this data.
# 
# 5. Create a Probability Plot of the HSDegree variable.
# 
# 6. Answer the following questions based on the Probability Plot:
#   
#   a. Based on what you see in this probability plot, is the distribution approximately normal? Explain how you know.
# 
# b. If not normal, is the distribution skewed? If so, in which direction? Explain how you know.
# 
# 7. Now that you have looked at this data visually for normality, you will now quantify normality with numbers using the stat.desc() function. Include a screen capture of the results produced.
# 
# 8. In several sentences provide an explanation of the result produced for skew, kurtosis, and z-scores. In addition, explain how a change in the sample size may change your explanation?
```
1. There are 8 elements in the dataset
  - Id              : Categorical, used as an identifier
  - Id2             : Categorical, used as an identifier
  - Geography       : Categorical
  - PopGroupID      : Categorical
  - RacesReported   : Quantitative
  - HSDegree        : Quantitative
  - BachDegree      : Quantitative

2. 
&nbsp;&nbsp;&nbsp;&nbsp;**head(survey_df)**

```{r head, echo=FALSE}
head(survey_df)
```
&nbsp;&nbsp;&nbsp;&nbsp;**str(survey_df)**
```{r str, echo=FALSE}
str(survey_df)
```
&nbsp;&nbsp;&nbsp;&nbsp;**nrow(survey_df)**
``` {r nrow, echo=FALSE}  
nrow(survey_df)
```
&nbsp;&nbsp;&nbsp;&nbsp;**ncol(survey_df)**
``` {r ncol, echo=FALSE}  
ncol(survey_df)
```
3. bins = 10. I used Rice's rule (cube root of the observations * 2) .



``` {r section 3 - Historgram, echo = FALSE}
#########################################################################
### Set bins using Rice's Rule - Cube Root of Observations squared
### Thanks to  Scott Breitbach for sharing this on the boards
#########################################################################
myBins <- as.integer((length(survey_df$HSDegree)^(1/3))*2)

x_label_plot <- "% High School Degree"
########################################
### Histogram, frequency Distribution
########################################
ggplot(survey_df, aes(HSDegree)) + 
  geom_histogram(
    aes(y=..density..), 
    bins = myBins, color = 'WHITE'
    )  + 
  theme_dark()  + 
  theme(
    panel.background = element_rect(fill = "linen")
    ) +
  labs(
    title= "Distribution of Population with a High School Degree",
    subtitle = paste("bins =",myBins),
    x=x_label_plot, 
    y="Count"
  )
```
4.

  - a. The Data appears to be modal. The measured values are continuous the degree of unimodality is directly related to the bins size. Once the bins setting is greater than 28, the data no longer appears to be unimodal. Sampling with a higher bins value doesn't seem to have significance since the range of values is fairly small.

  - b. The distribution does not appear to be symmetrical
  - c. The distribution does not appear to be bell shaped.
  - d. The distribution does not appear to be normal
  - e. The distribution appears to be negatively skewed.
  
  - f. 
  
``` {r section 4f, echo=FALSE}
################################################################################
### Combined Plot: Density Histogram, Normal Distribution, Probabilty Density
################################################################################
myBins <- as.integer((length(survey_df$HSDegree)^(1/3))*2)
ggplot(survey_df, aes(HSDegree)) +
  ###########
  #Theming 
  ###########
  theme_dark() + 
  theme(
    panel.background = element_rect(fill = "linen") 
  ) +
  ###########
  #Labels
  ###########
  labs(
    title= "Distribution of Population with a High School Degree",
    subtitle = paste( paste("bins =",myBins), ", Red - Normal Curve"),
    x=x_label_plot, 
    y="Distribution Density"
  )+
  ##################################
  # Base Layer: Density Histogram
  ##################################
  geom_histogram(
    aes(y=..density..), 
    bins = myBins, 
    color = 'WHITE'
  ) +
  #########################################
  # Add Line: Normal Distribution - Green
  #########################################
  stat_function(
    fun = dnorm, 
    args = list(mean = mean(survey_df$HSDegree), sd = sd(survey_df$HSDegree)),color="red", size=1, 
  ) 
```

  - g. The data does not appear to be normally distributed, therefore the normal distribution model cannot be used for this data.

5.

``` {r section 5, echo=FALSE}
################################################################################
### Combined Plot: Density Histogram, Normal Distribution, Probabilty Density
################################################################################
ggplot(survey_df, aes(HSDegree)) +
  ###########
  #Theming 
  ###########
  theme_dark() + 
  theme(
    panel.background = element_rect(fill = "linen") 
  ) +
  ###########
  #Labels
  ###########
  labs(
    title= "Distribution of Population with a High School Degree",
    subtitle = paste( paste("bins =",myBins), ", Green - Probability Distribution    Red - Normal Curve"),
    x=x_label_plot, 
    y="Distribution Density"
  )+
  ##################################
  # Base Layer: Density Histogram
  ##################################
  geom_histogram(
    aes(y=..density..), 
    bins = myBins, 
    color = 'WHITE'
  ) +
  #########################################
  # Add Line: Normal Distribution - Red
  #########################################
  stat_function(
    fun = dnorm, 
    args = list(mean = mean(survey_df$HSDegree), sd = sd(survey_df$HSDegree)),color="red", size=1, 
  ) +
  ############################################
  # Add Line: Probability Density - Green
  ############################################
  geom_density(color="green", size=1)

```


6.
  - a.The probability plot is not approximately normal. The shape of the probability plot has a sharper, more peaked curve compared to the normal curve. I plotted both and the differences are readily apparent.
  
  - b. The probability plot is negatively skewed because it trails off to the left. There isn't enough data to finish the plot to the right of the peak which adds to the appearance of negative skewness.

7.

``` {r section 7, echo=FALSE}
stat.desc(survey_df$HSDegree, basic=FALSE, norm=TRUE)
```
8. In a normal distribution the values for skew and kurtosis are 0. The further the values are from 0, the more likely the data is not normally distributed. The data has a kurtosis 4.352 and a skewness of -1.674. These values indicate the data is not normally distributed.

The data may be skewed because the sample size is not sufficiently large to represent a normal distribution according to the Central Limit Theorem. Increasing the sample size to 200 may result in a more normal distribution. 

Non-statistically speaking the data is measuring the frequency of a high school education across sub-section of the US population. the quality of education varies greatly by geography in the United States and is directly linked to individual prosperity or the lack thereof measured by inequality. Given socio-economic issues endemic to the United States I hypothesize that basic education levels are not normally distributed (in the statistical sense) as a whole.

I am unsure of how to interpret the z-scores in this context. I understand an individual z-scores can be used to indicate the probability of an individual score. But I'm unclear on how to use and interpret the z-scores as a whole. I plotted the z-scores below, but that didn't seem significant.

``` {r zScore, echo=FALSE}
zScores <- (survey_df$HSDegree - mean(survey_df$HSDegree)) / sd(survey_df$HSDegree)
ggplot(survey_df, aes(zScores)) + 
          geom_histogram(
            aes(), 
            bins = myBins, color = 'WHITE'
          )  + 
          theme_dark()  + 
          theme(
            panel.background = element_rect(fill = "linen")
          ) +
          labs(
            title= "Z-score Distribution",
            subtitle = paste("bins =",myBins),
            x="Z-Scores", 
            y="Frequency"
          )
```
